Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-09-14 supplement.tex
Update Suppl Fig 4
supplement: edits for consistency etc, move tables to end

%%% macros for notation in DP framework
\newcommand{\network}{\mathcal{N}}
\newcommand{\val}{\bar S} % valuation aka assignment
\newcommand{\dep}{\operatorname{dep}}
\newcommand{\energy}[1]{\operatorname{e}_{#1}}
\newcommand{\numberof}{\operatorname{\#}}
\newcommand{\partfun}[1]{\mathcal{Z}_{#1}}
\newcommand{\separator}[2]{\operatorname{sep}(#1,#2)}
\newcommand{\difference}[2]{\operatorname{diff}(#1 \rightarrow #2)}
\newcommand{\real}{\mathbb{R}}
\newcommand{\genmarg}[1]{(\!|\!#1\!|\!)}
\newcommand{\gencomb}[1]{\langle\!|#1|\!\rangle}
\newcommand{\Message}[2]{m_{#1\rightarrow #2}}

@@

\newcommand{\evalfor}[2]{#1\llbracket{}#2\rrbracket{}}
\newcommand{\substitute}[2]{#1\!\oplus\!#2}

@@

\section{Expressing higher arity functions as cliques in the dependency graph}
\label{appsec:dependency-cliques}

Our implementation relies on external tools for the tree decomposition. Therefore it is not trivial that the weight functions (i.e. terms of the energy function) can be captured within the tree decompositions returned by a specific tool. In other words, it is not immediately clear how one can construct a cluster tree from an arbitrary tree decomposition for a given network.

Crucially, one can show that, as long as the arguments $\dep(f)$ of a function $f$ form a clique within the network $\network$, then there exists at least one node in the tree decomposition that contains $\dep(f)$, thus allowing the evaluation of $f$.
%
While this result is certainly part of folklore, we emphasize the importance for the generality of this work and present a brief direct proof for convenience.

@@

\begin{theorem}[Correctness of Alg.~\ref{alg:pf}]
  \label{the:pfalgo-correctness}
  As computed by Alg.~\ref{alg:pf} for cluster tree $(T,\chi,\phi)$,
  the messages $\Message{u}{v}$, for all edges $u\to{}v\in T$, yield
  the partition functions of subtree of $u$ for the partial sequences
  $\val\in\partseqs(\separator{u}{v})$, i.e. the messages satisfy
  \begin{equation}
   \evalfor{\Message{u}{v}}{\val} = \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})} \quad
   \prod_{f\in\phi(T_r(u))} \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}),\label{eq:pfalgo-correct}
 \end{equation}
 where $\chi(T_r(u))$ denotes all $\chi$-assigned positions of nodes in $T_r(u)$;
 respectively $\phi(T_r(u))$; all $\phi$-assigned functions.
%
\end{theorem}

\begin{proof}
  Note that in more concise notation, Alg.~\ref{alg:pf} computes
  messages such that
\begin{equation}
  \evalfor{\Message{u}{v}}{\val} := \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}}) \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}.\label{eq:messages}
\end{equation}

Proof by induction on $T$. If $u$ is a leaf, $\chi(r) = \chi(T_r(u))$,
there are no messages sent to $u$, and $\phi(u) = \phi(T_r(u));$ implying
Eq.~$(\ref{eq:pfalgo-correct}).$
%
Otherwise, since the algorithm traverses edges in postorder, $u$ received from its children
$w_1,\dots,w_q$ the messages $\Message{w_1}{u}, \dots, \Message{w_q}{u}$, which satisfy Eq.~$(\ref{eq:pfalgo-correct})$ (induction hypothesis). Let $\val\in\partseqs(\separator{u}{v})$; then, $\evalfor{\Message{u}{v}}{\val}$ is computed by the algorithm according to Eq.~$(\ref{eq:messages})$. We rewrite as follows
\begin{align*}
  & \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
    \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}\\
  & = \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
    \prod_{i=1}^q \evalfor{\Message{w_i}{u}}{\substitute{\val'}{\val}}\\
  & =_{IH}
    \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
    \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
    \prod_{i=1}^q \sum_{\val''\in\partseqs(\chi(T_r(w_i))-\separator{w_i}{u})} \quad
    \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\substitute{\val'}{\val}}}) \\
& =_{*}
    \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
  \sum_{\val''\in\partseqs(\bigcup_{i=1}^q\chi(T_r(w_i))-\separator{w_i}{u})} \quad
  \prod_{f\in \phi(u) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})
  \prod_{f\in\phi(T_r(w_i))} \exp(-\beta \evalfor{f}{\substitute{\val''}{\val'} | \val}) \\
  & =_{(**)} \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})}\quad
    \prod_{f\in \phi(T_r(u)) } \exp(-\beta \evalfor{f}{\substitute{\val'}{\val}})\\
\end{align*}
To see (*) and (**), we observe:
\begin{itemize}
\item The sets $\chi(u)-\separator{u}{v}$ and
  $\chi(T_r(w_i))-\separator{w_i}{u}$ are all disjoint due to
  Def.~\ref{def:treedecomp}, property 2. First, this property implies
  that any shared position between the subtrees of $w_i$ and $w_j$
  must be in $\chi(w_i)$, $\chi(w_j)$ and $\chi(u)$, thus the
  positions of $\chi(T_r(w_i))-\separator{w_i}{u}$ are
  disjoint. Second, if a position $\chi(T_r(w_i))$ occurs in
  $\chi(u)$, it must occur in $\chi(w_i)$ and consequently in $\separator{u}{v}$.
\item The union of the sets $\chi(u)-\separator{u}{v}$ and
  $\chi(T_r(w_i))-\separator{w_i}{u}$ is $\chi(T_r(u))-\separator{u}{v}).$
\end{itemize}

@@

\begin{proposition}
  \label{prop:general-complexity}
Given a cluster tree $(T,\chi,\phi)$, $T=(V,E)$ of the RNA design network $\network=(\X,\F)$ with treewidth $w$ and maximum separator size $s$, computing the partition function by Algorithm~\ref{alg:pf} takes $O((|F|+|V|)\cdot 4^{w+1})$ time and $O(|V| 4^s)$ space (for storing all messages). The sampling step has time complexity of $O((|F|+|V|)\cdot 4^D)$ per sample.
\end{proposition}

@@

%\textcolor{red}{\textbf{SW: Simpler alternative to
%    Sec.~\ref{sec:first-improvedComplexity}. Note also that the use of
%    variables in Sec.~\ref{sec:first-improvedComplexity} conflicts
%    with the (recent) specialization of constraint networks to RNA design
%    networks.}}

@@

\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
    %{\sf \bfseries B}
    \includegraphicstop[width=0.8\textwidth]{Figs/energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies using multi-dimensional Boltzmann
    sampling. We show the Turner energy distributions for
    three target structures S1, S2 and S3. By
    targeting the respective free energies $(-40,-40,-20)$ and
    $(-20,-20,-20)$ for the target structures (annotated in red and
    blue, resepectively), we demonstrate the
    effectivity of our adaptive multi-dimensional Boltzmann sampling
    procedure; moreover, for comparison, the distributions for uniform
    (U) and Boltzmann (B) samples---respectively associated with
    homogenous weights $1$ and $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist}
\end{figure*}

@@

We first remark that the partial derivative of $\mathcal{B}$ yields
\begin{align*}
  \frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell} = -E_\ell(S)\cdot \pi_\ell^{-E_\ell(S)-1}\prod_{\substack{i=1\\i\neq \ell}}^{k} \pi_i^{-E_i(S)} = \frac{-E_\ell(S)}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S) = -E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }
\end{align*}
while the partial derivative of $\mathcal{Z}$ gives
\begin{align*}
  \frac{\partial \partfun{\pmb{\pi}}(S)}{\partial \pi_\ell} = \sum_{S'} -E_\ell(S')\cdot \mathbb{P}(S'\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }= -\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell}.
\end{align*}
From these expressions, we conclude that
\begin{align*}
  \frac{\partial \mathbb{E}(E_\ell\mid \pmb{\pi})}{\partial \pi_\ell} &= \sum_S E_\ell(S)\cdot\frac{\partial \mathbb{P}(S\mid \pmb{\pi})}{\partial \pi_\ell}\\
  %
  & = \sum_S E_\ell(S)\left(\frac{\frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell}}{\mathcal{Z}_{\pmb{\pi}}} - \frac{\frac{\partial \mathcal{Z}_{\pmb{\pi}}}{\partial \pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
  %
  & = \sum_S E_\ell(S)\left(\frac{-E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell }}{\mathcal{Z}_{\pmb{\pi}}} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
  %
  & = \sum_S - \frac{E_\ell(S)^2\cdot\mathbb{P}(S\mid \pmb{\pi})}{\pi_\ell} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})}{\pi_\ell}\sum_S E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\\
  & = -\frac{\mathbb{E}(E_\ell^2\mid \pmb{\pi}) - \mathbb{E}(E_\ell\mid \pmb{\pi})^2}{\pi_\ell}
\end{align*}
In the numerator of the above expression, one recognizes the variance of the distribution of $E_\ell$.
Remark that a variance is always non-negative and, in our case, also strictly positive for $\pmb{\pi}$, $\pi_\ell>0$, as soon as there exist at least two distinct sequences $S$ and $S'$ such that $E_\ell(S)\neq E_\ell(S')$.
Moreover, weights are always positive so the partial derivative in $\pi_\ell$ is always positive. Ultimately, it is always possible to increase (resp. decrease) the expected free energy of a structure by simply decreasing (resp. increasing) its weight, supporting our weight optimization procedure.
\newpage
\section{Detailed result of MultiDefect analysis}\label{sec:validity}
N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.
\input{table}

@@

>>>>

%%% macros for notation in DP framework
\newcommand{\val}{\bar S} % valuation aka assignment
\newcommand{\dep}{\operatorname{dep}}
\newcommand{\energy}[1]{\operatorname{e}_{#1}}
\newcommand{\numberof}{\operatorname{\#}}
\newcommand{\partfun}[1]{\mathcal{Z}_{#1}}
\newcommand{\separator}[2]{\operatorname{sep}(#1,#2)}
\newcommand{\difference}[2]{\operatorname{diff}(#1 \rightarrow #2)}
\newcommand{\real}{\mathbb{R}}
\newcommand{\genmarg}[1]{(\!|\!#1\!|\!)}
\newcommand{\gencomb}[1]{\langle\!|#1|\!\rangle}
\newcommand{\Message}[2]{m_{#1\rightarrow #2}}

@@

\newcommand{\evalfor}[2]{#1(#2)}
\newcommand{\substitute}[2]{#1\cup#2}

@@

\section{Tree decomposition with hyper-dependencies}
\label{appsec:dependency-cliques}

Our implementation relies on external tools for the tree decomposition, which do not directly support tree decomposition of hyper-graphs. Therefore it is not immediately obvious how 
to obtain a suitable tree decomposition, such that even multi-ary contributions are respected and a cluster tree can be generated.

Crucially, one can show that, as long as certain vertices of a graph $G$ from a clique, there exists at least one node in the tree decomposition of $G$ that contains all such vertices.
%
This allows us to generate suitable tree decompositions, which respect the hyper-edges of our dependency graph, by expanding the hyper-edges to cliques, and then calling the tree-decomposer on the resulting (non-hyper-)graph.

%
While the result is certainly part of folklore, we emphasize the importance for the generality of this work and present a brief direct proof for convenience.

@@

    \begin{theorem}[Correctness of Alg.~\ref{alg:pf}]
      \label{the:pfalgo-correctness}
      As computed by Alg.~\ref{alg:pf} for cluster tree $(T,\chi,\phi)$,
      the messages $\Message{u}{v}$, for all edges $u\to{}v\in T$, yield
      the partition functions of subtree of $u$ for the partial sequences
      $\val\in\partseqs(\separator{u}{v})$, i.e. the messages satisfy
      \begin{equation}
       \evalfor{\Message{u}{v}}{\val} = \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})} \quad
       \prod_{f\in\phi(T_r(u))} \exp(-\evalfor{f}{\substitute{\val'}{\val}}),\label{eq:pfalgo-correct}
     \end{equation}
     where $\chi(T_r(u))$ denotes all $\chi$-assigned positions of nodes in $T_r(u)$;
     respectively $\phi(T_r(u))$; all $\phi$-assigned functions.
    %
    \end{theorem}

    \begin{proof}
      Note that in more concise notation, Alg.~\ref{alg:pf} computes
      messages such that
    \begin{equation}
      \evalfor{\Message{u}{v}}{\val} := \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
      \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}}) \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}.\label{eq:messages}
    \end{equation}

    Proof by induction on $T$. If $u$ is a leaf, $\chi(r) = \chi(T_r(u))$,
    there are no messages sent to $u$, and $\phi(u) = \phi(T_r(u));$ implying
    Eq.~$(\ref{eq:pfalgo-correct}).$
    %
    Otherwise, since the algorithm traverses edges in postorder, $u$ received from its children
    $w_1,\dots,w_q$ the messages $\Message{w_1}{u}, \dots, \Message{w_q}{u}$, which satisfy Eq.~$(\ref{eq:pfalgo-correct})$ (induction hypothesis). Let $\val\in\partseqs(\separator{u}{v})$; then, $\evalfor{\Message{u}{v}}{\val}$ is computed by the algorithm according to Eq.~$(\ref{eq:messages})$. We rewrite as follows
    \begin{align*}
      & \sum_{\val'\in\partseqs(\difference{u}{v})}\quad
        \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})
        \prod_{(w\to{}u) \in T} \evalfor{\Message{w}{u}}{\substitute{\val'}{\val}}\\
      & = \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
        \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})
        \prod_{i=1}^q \evalfor{\Message{w_i}{u}}{\substitute{\val'}{\val}}\\
      & =_{IH}
        \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
        \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})
        \prod_{i=1}^q \sum_{\val''\in\partseqs(\chi(T_r(w_i))-\separator{w_i}{u})} \quad
        \prod_{f\in\phi(T_r(w_i))} \exp(-\evalfor{f}{\substitute{\val''}{\substitute{\val'}{\val}}}) \\
    & =_{*}
        \sum_{\val'\in\partseqs(\chi(u)-\separator{u}{v})}\quad
      \sum_{\val''\in\partseqs(\bigcup_{i=1}^q\chi(T_r(w_i))-\separator{w_i}{u})} \quad
      \prod_{f\in \phi(u) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})
      \prod_{f\in\phi(T_r(w_i))} \exp(-\evalfor{f}{\substitute{\val''}{\val'} \cup \val}) \\
      & =_{(**)} \sum_{\val'\in\partseqs(\chi(T_r(u))-\separator{u}{v})}\quad
        \prod_{f\in \phi(T_r(u)) } \exp(-\evalfor{f}{\substitute{\val'}{\val}})\\
    \end{align*}
    To see (*) and (**), we observe:
    \begin{itemize}
    \item The sets $\chi(u)-\separator{u}{v}$ and
      $\chi(T_r(w_i))-\separator{w_i}{u}$ are all disjoint due to
      Def.~\ref{def:treedecomp}, property 2. First, this property implies
      that any shared position between the subtrees of $w_i$ and $w_j$
      must be in $\chi(w_i)$, $\chi(w_j)$ and $\chi(u)$, thus the
      positions of $\chi(T_r(w_i))-\separator{w_i}{u}$ are
      disjoint. Second, if a position $\chi(T_r(w_i))$ occurs in
      $\chi(u)$, it must occur in $\chi(w_i)$ and consequently in $\separator{u}{v}$.
    \item The union of the sets $\chi(u)-\separator{u}{v}$ and
      $\chi(T_r(w_i))-\separator{w_i}{u}$ is $\chi(T_r(u))-\separator{u}{v}).$
    \end{itemize}

@@

    \begin{proposition}
      \label{prop:general-complexity}
    Given a cluster tree $(T,\chi,\phi)$, $T=(V,E)$ of the set of contributions $\F{}$ with treewidth $w$ and maximum separator size $s$, computing the partition function by Algorithm~\ref{alg:pf} takes $O((|F|+|V|)\cdot 4^{w+1})$ time and $O(|V| 4^s)$ space (for storing all messages). The sampling step has time complexity of $O((|F|+|V|)\cdot 4^D)$ per sample.
    \end{proposition}

@@

@@

    \begin{figure*}[t]
      \begin{center}
        %{\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
        %{\sf \bfseries B}
        \includegraphicstop[width=0.8\textwidth]{Figs/energy_distribution}
      \end{center}
      \caption{%
        Targeting specific energies using multi-dimensional Boltzmann
        sampling. We show the Turner energy distributions for
        three target structures R1, R2 and R3. By
        targeting the respective free energies $(-40,-40,-20)$ and
        $(-20,-20,-20)$ for the target structures (annotated in red and
        blue, resepectively), we demonstrate the
        effectivity of our adaptive multi-dimensional Boltzmann sampling
        procedure; moreover, for comparison, the distributions for uniform
        (U) and Boltzmann (B) samples---respectively associated with
        homogenous weights $1$ and $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist}
\end{figure*}

@@

We first remark that the partial derivative of $\mathcal{B}$ yields
\begin{align*}
  \frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell} = -E_\ell(S)\cdot \pi_\ell^{-E_\ell(S)-1}\prod_{\substack{i=1\\i\neq \ell}}^{k} \pi_i^{-E_i(S)} = \frac{-E_\ell(S)}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S) = -E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }
\end{align*}
while the partial derivative of $\mathcal{Z}$ gives
\begin{align*}
  \frac{\partial \partfun{\pmb{\pi}}(S)}{\partial \pi_\ell} = \sum_{S'} -E_\ell(S')\cdot \mathbb{P}(S'\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell }= -\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\partfun{\pmb{\pi}}}{\pi_\ell}.
\end{align*}
From these expressions, we conclude that
\begin{align*}
  \frac{\partial \mathbb{E}(E_\ell\mid \pmb{\pi})}{\partial \pi_\ell} &= \sum_S E_\ell(S)\cdot\frac{\partial \mathbb{P}(S\mid \pmb{\pi})}{\partial \pi_\ell}\\
  %
  & = \sum_S E_\ell(S)\left(\frac{\frac{\partial \mathcal{B}_{\pmb{\pi}}(S)}{\partial \pi_\ell}}{\mathcal{Z}_{\pmb{\pi}}} - \frac{\frac{\partial \mathcal{Z}_{\pmb{\pi}}}{\partial \pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
  %
  & = \sum_S E_\ell(S)\left(\frac{-E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell }}{\mathcal{Z}_{\pmb{\pi}}} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})\cdot \frac{\mathcal{Z}_{\pmb{\pi}}}{\pi_\ell}\cdot\mathcal{B}_{\pmb{\pi}}(S)}{\mathcal{Z}_{\pmb{\pi}}^2}\right)\\
  %
  & = \sum_S - \frac{E_\ell(S)^2\cdot\mathbb{P}(S\mid \pmb{\pi})}{\pi_\ell} + \frac{\mathbb{E}(E_\ell\mid \pmb{\pi})}{\pi_\ell}\sum_S E_\ell(S)\cdot \mathbb{P}(S\mid \pmb{\pi})\\
  & = -\frac{\mathbb{E}(E_\ell^2\mid \pmb{\pi}) - \mathbb{E}(E_\ell\mid \pmb{\pi})^2}{\pi_\ell}
\end{align*}
In the numerator of the above expression, one recognizes the variance of the distribution of $E_\ell$.
Remark that a variance is always non-negative and, in our case, also strictly positive for $\pmb{\pi}$, $\pi_\ell>0$, as soon as there exist at least two distinct sequences $S$ and $S'$ such that $E_\ell(S)\neq E_\ell(S')$.
Moreover, weights are always positive so the partial derivative in $\pi_\ell$ is always positive. Ultimately, it is always possible to increase (resp. decrease) the expected free energy of a structure by simply decreasing (resp. increasing) its weight, supporting our weight optimization procedure.

@@

\section{Detailed result of MultiDefect analysis}\label{sec:validity}
N/A values correspond to data that could not be obtained by the time of this submission for two main reasons: In the case of the initial sampling (left columns), they correspond to instances which, in conjunction with an expressive energy model, resulted in very high tree-width, leading to unreasonable memory requirements incompatible with our available computing resources; The case of missing values after optimization  (right columns), they indicate situations where the initial optimization took too long ($\approx$ 1 day) and was killed.
\input{table}
